Lock-free Skiplist-based Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-11 paper/paper.tex
Shavit & Lotan & Herlihy

A crucial observation is that nodes nodes which are only partially connected
do not affect correctness of the data structure. As soon as the first level (i.e. \lstinline|node.level[0]|)
has been successfully connected, a node is considered to be in the SkipList.
Therefore, both insertions and deletions can proceed in steps - insertions proceed bottom-up while
deletions proceed top-down, holding locks only for the current level.

\citeauthor{herlihy2012art} \cite{herlihy2012art} later described a variant of this idea 
that is lock-free and quiescently consistent (but not linearizable).

>>>>

A crucial observation is that nodes nodes which are only partially connected
do not affect correctness of the data structure. As soon as the first level (i.e. \lstinline|node.level[0]|)
has been successfully connected, a node is considered to be in the SkipList.
Therefore, both insertions and deletions can be split into steps --- insertions
proceed bottom-up while deletions proceed top-down, holding locks only for the current level.

\begin{lstlisting}
struct node_t {
  [...] /**< Standard node members as above. */
  atomic<bool> deleted; /**< Initially false. */
  time_t timestamp;
  lock_t locks[level + 1];
};
\end{lstlisting}

Likewise, deletions are split into a logical phase, atomically setting the \lstinline|node.deleted|
flag; and a physical phase performs the actual pointer manipulations and can be seen as a simple
call to the underlying SkipList's \lstinline|sl_delete| function.

A \lstinline|DeleteMin| call starts the list head, and attempts to atomically set
the deletion flag using a \lstinline|CAS(node.deleted, false, true)| call (or equivalent constructs).
If it succeeds, the current node is physically deleted and returned to the caller. Otherwise,
\lstinline|node.next[0]| is set as the new current node and the procedure is repeated.
If the end of the list is reached, \lstinline|DeleteMin| returns false to indicate an empty list.

Note that so far, this implementation is not linearizable: consider the case in which a
slow thread A is in the middle of a \lstinline|DeleteMin| call. Within this context, we refer to 
the node with key $i$ as ``node $i$'', or simply $i$. Several \lstinline|CAS|
operations have failed, and A is currently at node $j$.
A fast thread B then first inserts a node $i$, followed by a node $k$ such that
$i < j < k$, i.e. the former and latter nodes are inserted, respectively, before and after
thread A's current node. Assuming further that all nodes between $j$ and $k$ have already
been deleted, then thread A will return node $k$. This execution is not linearizable; however,
it is quiescently consistent since operations can be reordered to correspond to some sequential
execution at quiescent periods.

\citeauthor{shavit2000skiplist} counteract this by introducing a \lstinline|timestamp| for each
node which is set upon successful insertion. Each \lstinline|DeleteMin| operation 
simply ignores all nodes it sees that have not been fully inserted at the time it was called.

Explicit memory management is required to avoid dereferencing pointers to freed memory areas
by other threads after physical deletion. This implementation uses a dedicated garbage collector
thread in combination with a timestamping mechanism which frees \lstinline|node|'s memory only
when all threads that might have seen a pointer to \lstinline|node| have exited the data structure.

\citeauthor{herlihy2012art} \cite{herlihy2012art} later described and implemented a variant of this idea 
in Java that is lock-free and quiescently consistent. While mostly identical, notable differences are
that a) the new variant is based on a lock-free skiplist, b) the timestamping mechanism was not
employed and thus linearizability was lost, and c) explicit memory management is not required
because the Java virtual machine provides a garbage collector.
